---
title: "Question 03 part i"
output: html_document
date: "2024-04-02"
---

```{r}
# importing a necessary library
library(dplyr)
```

```{r}
# Setting the working directory 
setwd("D:/User/University/Year 2/Programming for Data Science/Programming Coursework/R cleaned files")
data_2007 <- read.csv("data_2007.csv")
```

```{r}
# Setting the working directory 
setwd("D:/User/University/Year 2/Programming for Data Science/Programming Coursework/R cleaned files")
airport_data <- read.csv("airport_data.csv")
```

```{r}
# viewing the first few records of the data set
head(data_2007)
```

```{r}
# dependent variable is categorical
divert_counts <- table(data_2007$Diverted); divert_counts
```
# Imputation of the null values
```{r}
colSums(is.na(data_2007))
```

```{r}
columns_with_null_values <- colnames(data_2007)[apply(data_2007, 2, anyNA)]
print(columns_with_null_values)
```
# in R, the column 'TailNum' for any presence of missing values is not returned, but after looking they are empty cells. 
# However, the distribution is not checked since it does have any impact to the model
```{r}
# Columns with missing values
columns <- c('DepTime', 'ArrTime', 'ActualElapsedTime', 'CRSElapsedTime', 'AirTime', 'ArrDelay', 'DepDelay')

# Iterating over the columns containing null values to check each variable's distribution
for (column in columns) {
  # Checking to make sure that the null values were not removed accidently
  if (anyNA(data_2007[[column]])) {
    # Histogram for column
    hist(data_2007[[column]], main = paste('Distribution of', column), xlab = 'Values', ylab = 'Frequency', breaks = 100)
  }
}
```
#the variables resembling uniform distibution will be imputed using the mean and the skewed ones with median the column 'TailNum' will be ignored as it is a alphanumeric code used for identification
```{r}
# data cleaning for missing values through imputation

# uniformly distributed variables
data_2007$ArrTime[is.na(data_2007$ArrTime)] <- mean(data_2007$ArrTime, na.rm = TRUE)
data_2007$DepTime[is.na(data_2007$DepTime)] <- mean(data_2007$DepTime, na.rm = TRUE)

# skewed distributions
data_2007$ActualElapsedTime[is.na(data_2007$ActualElapsedTime)] <- median(data_2007$ActualElapsedTime, na.rm = TRUE)
data_2007$CRSElapsedTime[is.na(data_2007$CRSElapsedTime)] <- median(data_2007$CRSElapsedTime, na.rm = TRUE)
data_2007$AirTime[is.na(data_2007$AirTime)] <- median(data_2007$AirTime, na.rm = TRUE)
data_2007$ArrDelay[is.na(data_2007$ArrDelay)] <- median(data_2007$ArrDelay, na.rm = TRUE)
data_2007$DepDelay[is.na(data_2007$DepDelay)] <- median(data_2007$DepDelay, na.rm = TRUE)
```

```{r}
# rechecking to confirm if the missing values have been imputed
colSums(is.na(data_2007))
```

```{r}
names(data_2007)
```

```{r}
filtered_data_2007 <- data_2007[, c('Diverted', 'Month', 'DayofMonth', 'DayOfWeek', 'CRSDepTime',
                                     'CRSArrTime', 'DepDelay', 'UniqueCarrier', 'Origin', 'Dest', 'Distance', 'Cancelled', 
                                     'CarrierDelay', 'WeatherDelay', 'NASDelay', 'SecurityDelay', 'LateAircraftDelay', 'TailNum')]
```

```{r}
# a view of the first few records of the airport data set
head(airport_data)
# column names of the airport data set
names(airport_data)
```
# it was found by observation that the column 'iata' in the airport dataset is similar to the Origin and Dest columns in the filtered 2007 dataset. further information about coordinates can be extracted

# Merging the datasets based on appropriate columns
```{r}
# renaming the column 'Dest' in filtered_data_2007 to 'iata'
names(filtered_data_2007)[names(filtered_data_2007) == 'Dest'] <- 'iata'

# Print the first row of the modified data frame to check the column names
print(head(filtered_data_2007, 1))
```

```{r}
# combining the airport data set and the filtered_data_2007 set
# using 'iata' and 'Origin' columns to merge to info about destination coordinates
merged_set_v1 <- merge(filtered_data_2007, airport_data[, c('iata', 'lat', 'long')], by = 'iata', all.x = TRUE)
print(head(merged_set_v1))
```

```{r}
# renaming the columns lat and long to dest lat and long to get the specific coordinates
colnames(merged_set_v1)[which(colnames(merged_set_v1) == 'lat')] <- 'Dest lat'
colnames(merged_set_v1)[which(colnames(merged_set_v1) == 'long')] <- 'Dest long'

print(head(merged_set_v1))
```
# remerging the datasets using 'iata' column to get the origin coordinates
```{r}
# Removing the 'iata' column from the merged data frame

merged_set_v1 <- subset(merged_set_v1, select = -c(iata))
print(head(merged_set_v1))

```

```{r}
# renaming the column 'Origin' in filtered_data_2007 to 'iata'
colnames(merged_set_v1)[colnames(merged_set_v1) == 'Origin'] <- 'iata'
print(head(merged_set_v1))
```

```{r}
# combining the airport data set and the merged_set_v1

# re-merging with airport data set
airport_columns <- c('iata', 'airport', 'lat', 'long')
merged_set_v2 <- merge(merged_set_v1, airport_data[, airport_columns], by = 'iata', all.x = TRUE)
head(merged_set_v2)
```

```{r}
# renaming the columns lat and long to dest lat and long to get the specific coordinates
merged_set_v2 <- merged_set_v2 %>% rename('Origin lat' = lat, 'Origin long' = long)

# Removing the 'iata' column from the merged DataFrame
merged_set_v2 <- merged_set_v2 %>% select(-iata)
```

```{r}
# Encoding the Categorical variable, 'UniqueCarrier'
table(merged_set_v2$UniqueCarrier)
```
# Encoding categorical variables
```{r}
# Setting the data type to factor before ENCODING the column
merged_set_v2$UniqueCarrier <- factor(merged_set_v2$UniqueCarrier)

merged_set_v2$UniqueCarrier_Encoded <- as.integer(merged_set_v2$UniqueCarrier)

# Removing the 'UniqueCarrier' column
merged_set_v3 <- merged_set_v2[, -which(names(merged_set_v2) == 'UniqueCarrier')]
print(merged_set_v3)
```

```{r}
# Encoding 'airport' column datatype to factor
merged_set_v2$airport <- factor(merged_set_v2$airport)
merged_set_v2$airport_Encoded <- as.integer(merged_set_v2$airport)

# removing the airport column
merged_set_v3 <- merged_set_v2[, -which(names(merged_set_v2) == 'airport')]
# removing the Uniquecarrier column
merged_set_v3 <- merged_set_v3[, -which(names(merged_set_v3) == 'UniqueCarrier')]
print(merged_set_v3)
```

```{r}
# encoding the date attributes
# Setting the data type to factor and ENCODING the resulting column 

merged_set_v3$Month_Encoded <- as.integer(as.factor(merged_set_v3$Month))
merged_set_v3$DayofMonth_Encoded <- as.integer(as.factor(merged_set_v3$DayofMonth))
merged_set_v3$DayOfWeek_Encoded <- as.integer(as.factor(merged_set_v3$DayOfWeek))

merged_set_v4 <- merged_set_v3[, !(names(merged_set_v3) %in% c('Month', 'DayofMonth', 'DayOfWeek'))]

head(merged_set_v4)
```

```{r}
# converting the variables datatype to factor
merged_set_v4$Diverted <- as.integer(as.factor(merged_set_v2$Diverted))
merged_set_v4$Cancelled <- as.integer(as.factor(merged_set_v2$Cancelled))
```

```{r}
# getting the necessary columns for the model

columns_for_model <- merged_set_v4[c('Diverted', 'CRSDepTime', 'CRSArrTime', 'Distance', 'Cancelled', 'CarrierDelay', 'DepDelay', 
                                     'WeatherDelay', 'NASDelay', 'SecurityDelay', 'LateAircraftDelay', 'Origin lat', 'Origin long', 
                                     'Dest lat', 'Dest long', 'UniqueCarrier_Encoded', 'airport_Encoded', 'Month_Encoded', 'DayOfWeek_Encoded')]
```

```{r}
# saving the file
write.csv(columns_for_model, file = "D:/User/University/Year 2/Programming for Data Science/Programming Coursework/R files/columns_for_model.csv", row.names = FALSE)
```
